---
title: "Example Report"
author: "Magdalena A. Tkacz"
subtitle: "dataset: <to-do>"
output:
  bookdown::html_document2:
    df_print: paged
    theme: spacelab
    highlight: zenburn
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/"
)
```

<div style="
background-color:#e6d7eb;
border-left:8px solid #5c0775;
padding:12px 16px;
margin:16px 0;
color:#111111;
">

This document provides an example report structure based on the CRISP-DM methodology. 

Students may use, modify, or extend the template as appropriate for their project.

The practical examples presented in this report are based on `penguin_mess`, a modified version of the `palmerpenguins` dataset created for educational purposes.

The dataset was intentionally altered to include selected data quality issues and should not be regarded as the original `palmerpenguins` dataset.


</div>



# Assessment Summary

Student name: Pen Guin

Intended grade:☐ 3.0   ☐ 3.5   ☑ 4.0   ☐ 4.5   ☐ 5.0

# Student Declaration

By submitting this report, I declare that:

- I understand all parts of the submitted work;
- I am able to explain and justify the analytical decisions presented in the report;
- the self-assessment table has been completed honestly;
- any use of AI tools was limited to assistance and does not substitute for my own understanding of the analysis;
- all external sources and tools used during the project have been properly acknowledged.

Submission of this report constitutes acceptance of this declaration.

# 0. Executive Summary

## Report Summary
This report presents a complete Data Mining workflow based on the CRISP-DM methodology. The analysis uses the penguin_mess dataset, a modified version of the Palmer Penguins dataset prepared for educational purposes.

The primary objective of the project is to predict penguin species using morphological measurements and selected categorical attributes. 

The analysis follows the CRISP-DM (Cross-Industry Standard Process for Data Mining) methodology, progressing through business understanding, data understanding, data preparation, exploratory analysis, modeling, and evaluation.


Two classification models were developed and compared. The results demonstrate the importance of proper data preparation and feature selection when building predictive models.

## Success Criteria

The project will be considered successful if:

- data quality issues are identified and addressed;
- meaningful exploratory analysis is performed;
- at least two classification models are developed;
- model performance is evaluated and compared;
- conclusions are supported by analytical evidence.

# 1. Business Understanding

The objective of this project is to develop a predictive model capable of identifying penguin species based on available measurements and characteristics.

Although the dataset is educational in nature, the problem reflects a common Data Mining scenario in which historical observations are used to classify new cases. The project also aims to demonstrate the complete CRISP-DM process rather than maximize predictive performance.

The project aims to answer the following question: 

> Can penguin species be accurately predicted using the available measurements and characteristics?

## Project Scope 

The analysis includes: 
- data understanding and quality assessment; 
- data preparation and cleaning; 
- exploratory data analysis (EDA); 
- development of classification models; 
- model evaluation and comparison; 
- interpretation of results. 

The project does not aim to build a production-ready system. Instead, it serves as a demonstration of the CRISP-DM process and fundamental Data Mining techniques.

## Problem Statement

Given a set of penguin observations containing morphological measurements and categorical characteristics, the objective is to develop a predictive model capable of correctly identifying the species of each penguin.

The project investigates which variables are most informative for species classification and evaluates whether reliable predictions can be obtained from the available data.

## Objective

The objective of this project is to investigate whether penguin species can be accurately predicted using available morphological measurements and categorical attributes. 

To achieve this goal, the dataset will be explored, cleaned, and prepared for modeling. Multiple classification models will be developed and compared in order to identify the most suitable approach for this prediction task.


## Expected Outcome

The expected outcome of this project is the development of a classification model capable of predicting penguin species based on available measurements and categorical attributes.
It is expected that the analysis will identify variables that are most informative for species classification and reveal patterns that distinguish different species.

Furthermore, the project is expected to demonstrate the impact of data preparation and feature selection on model performance, as well as provide a comparison of alternative modeling approaches.


# 2. Data Understanding

The purpose of this phase is to become familiar with the dataset and assess its suitability for the intended analysis. 
This includes examining the structure of the data, identifying variable types, evaluating data quality, and detecting potential issues that may affect subsequent stages of the project. 

Particular attention is given to missing values, inconsistencies, unusual observations, and the distribution of individual variables. Understanding these characteristics is essential before proceeding to data preparation and model development.

## Dataset Description

Provide a general description of the dataset.

## Variables

Describe the variables used in the analysis.

## Initial Exploration

```{r}
# Initial exploration
```

## Initial Observations

Describe your initial observations.

# 3. Data Preparation

## Missing Values

Describe how missing values were identified and handled.

```{r}
# Missing values analysis
```

## Data Cleaning

Describe any cleaning operations performed.

## Transformations

Describe any transformations applied to the data.

## Feature Selection / Engineering

Describe any variables removed, added, or modified.

# 4. Exploratory Data Analysis (EDA)

## Distribution Analysis

```{r}
# Distribution analysis
```

### Interpretation

Describe your findings.

## Relationship Analysis

```{r}
# Relationship analysis
```

### Interpretation

Describe your findings.

## Correlation Analysis

```{r}
# Correlation analysis
```

### Interpretation

Describe your findings.

## Outlier Analysis

```{r}
# Outlier analysis
```

### Interpretation

Describe your findings.

## Summary of EDA

Summarize the most important observations.

# 5. Modeling

## Model 1

### Model Description

Explain why this model was selected.

### Implementation

```{r}
# Model implementation
```

### Results

```{r}
# Model evaluation
```

### Interpretation

Interpret the results.

## Model 2

### Model Description

Explain why this model was selected.

### Implementation

```{r}
# Model implementation
```

### Results

```{r}
# Model evaluation
```

### Interpretation

Interpret the results.

# 6. Evaluation

## Model Comparison

Provide a comparison of the evaluated models.

| Metric | Model 1 | Model 2 |
|---------|---------|---------|
| | | |
| | | |
| | | |

## Discussion

Discuss the strengths and weaknesses of the models.

## Limitations

Discuss any limitations of the data, methodology, or results.

# 7. Recommendations

Provide recommendations based on the results of the analysis.

# 8. Conclusions

## Summary

Summarize the project.

## Lessons Learned

Describe what was learned during the analysis.

## Future Work

Suggest possible future improvements or extensions.

# References

List all external sources used.

# Appendix (Optional)

Additional tables, figures, code snippets, or supplementary material.