---
title: "Example Report"
author: "Magdalena A. Tkacz"
subtitle: "dataset: <to-do>"
output:
  html_document:
    df_print: paged
    theme: spacelab
    highlight: zenburn
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div style="
background-color:#e6d7eb;
border-left:8px solid #5c0775;
padding:12px 16px;
margin:16px 0;
color:#111111;
">

This document provides an example report structure based on the CRISP-DM methodology. 

Students may use, modify, or extend the template as appropriate for their project.

</div>

# Assessment Summary

Student name: ______________________

Intended grade: ☐ 3.0 ☐ 3.5 ☐ 4.0 ☐ 4.5 ☐ 5.0

# Student Declaration

By submitting this report, I declare that:

- I understand all parts of the submitted work;
- I am able to explain and justify the analytical decisions presented in the report;
- the self-assessment table has been completed honestly;
- any use of AI tools was limited to assistance and does not substitute for my own understanding of the analysis;
- all external sources and tools used during the project have been properly acknowledged.

Submission of this report constitutes acceptance of this declaration.

# 1. Business Understanding

## Problem Statement

Describe the problem addressed in the project.

## Objective

Define the objective of the analysis.

## Expected Outcome

Describe the expected outcome and its potential usefulness.

# 2. Data Understanding

## Dataset Description

Provide a general description of the dataset.

## Variables

Describe the variables used in the analysis.

## Initial Exploration

```{r}
# Initial exploration
```

## Initial Observations

Describe your initial observations.

# 3. Data Preparation

## Missing Values

Describe how missing values were identified and handled.

```{r}
# Missing values analysis
```

## Data Cleaning

Describe any cleaning operations performed.

## Transformations

Describe any transformations applied to the data.

## Feature Selection / Engineering

Describe any variables removed, added, or modified.

# 4. Exploratory Data Analysis (EDA)

## Distribution Analysis

```{r}
# Distribution analysis
```

### Interpretation

Describe your findings.

## Relationship Analysis

```{r}
# Relationship analysis
```

### Interpretation

Describe your findings.

## Correlation Analysis

```{r}
# Correlation analysis
```

### Interpretation

Describe your findings.

## Outlier Analysis

```{r}
# Outlier analysis
```

### Interpretation

Describe your findings.

## Summary of EDA

Summarize the most important observations.

# 5. Modeling

## Model 1

### Model Description

Explain why this model was selected.

### Implementation

```{r}
# Model implementation
```

### Results

```{r}
# Model evaluation
```

### Interpretation

Interpret the results.

## Model 2

### Model Description

Explain why this model was selected.

### Implementation

```{r}
# Model implementation
```

### Results

```{r}
# Model evaluation
```

### Interpretation

Interpret the results.

# 6. Evaluation

## Model Comparison

Provide a comparison of the evaluated models.

| Metric | Model 1 | Model 2 |
|---------|---------|---------|
| | | |
| | | |
| | | |

## Discussion

Discuss the strengths and weaknesses of the models.

## Limitations

Discuss any limitations of the data, methodology, or results.

# 7. Recommendations

Provide recommendations based on the results of the analysis.

# 8. Conclusions

## Summary

Summarize the project.

## Lessons Learned

Describe what was learned during the analysis.

## Future Work

Suggest possible future improvements or extensions.

# References

List all external sources used.

# Appendix (Optional)

Additional tables, figures, code snippets, or supplementary material.